Symbolic Abstract Data Type Inference

### 2015-11-02 discussion.tex
Corrections from Reviewer B.
Paragraph formatting.

First, while our theoretical foundation does cover “concurrency-aware” ADT
specifications~\cite{conf/podc/HemedR14}, i.e.,~non-atomic ADTs like the
\emph{rendezvous synchronizer}, \emph{barrier}, and \emph{exchanger}, which
have bounded width greater than $1$, our current implementation only
handles atomic ADTs. In principle, this limitation is not fundamental. The key
difference is in the enumeration ADT histories. For a given ADT,
determining whether a given history is admitted or not reduces
to checking whether there exists an execution of which the history is an
abstraction. For atomic ADTs, only a single sequential execution need be
examined, in which no two operations overlap. For non-atomic ADTs, every
possible interleaving of the internal actions of operations need be examined.
While we expect this calculation to remain feasible given that ADT complements
are normally represented with histories with few operations, we do expect it to
incur a noticeable cost.

Second, ADTs whose specifications rely on relations (besides equality) on
method argument and return values cannot be expressed with our notion of
histories based on matching schemes. A notable example of such an ADT is the
\emph{priority queue}, whose \emph{dequeue} operations return the
smallest/largest \emph{enqueue}d value which has not yet been dequeued.
%It appears that such ADTs are rare in practice.
Overcoming this limitation would
require a refinement to matching schemes which can track additional relations
among data values.

Finally, in some cases it is unclear how to express matching schemes
deterministically. For example, each \emph{wait} operation of \emph{semaphore}
ADT should naturally match the \emph{notify} operation which enabled it.
However, it is not possible to determine this based on operation labels alone,
and it is not clear whether all implementations effectively keep track of this
correspondence. Furthermore, choosing matches arbitrarily would be unsound, and
result in inferring ADTs with artificial constraints, e.g.,~that the notify and
wait operations behave according to a FIFO discipline, each wait matching the
oldest unmatched notify. Overcoming this limitation may require introducing a
notion of nondeterminism into matching schemes.

>>>>

First, while our theoretical foundation does cover “concurrency-aware” ADT
specifications~\cite{conf/podc/HemedR14}, i.e.,~non-atomic ADTs like the
\emph{rendezvous synchronizer}, \emph{barrier}, and \emph{exchanger}, which
have bounded width greater than $1$, our current implementation only
handles atomic ADTs. In principle, this limitation is not fundamental. The key
difference is in the enumeration of ADT histories. For a given ADT,
determining whether a given history is admitted or not reduces
to checking whether there exists an execution of which the history is an
abstraction. For atomic ADTs, only a single sequential execution need be
examined, in which no two operations overlap. For non-atomic ADTs, every
possible interleaving of the internal actions of operations need be examined.
While we expect this calculation to remain feasible given that ADT complements
are normally represented with histories with few operations, we do expect it to
incur a noticeable cost.

Second, ADTs whose specifications rely on relations (besides equality) on
method argument and return values cannot be expressed with our matching-scheme
based notion of histories. A notable example is the
\emph{priority queue}, whose \emph{dequeue} operations return the
smallest/largest \emph{enqueue}d value which has not yet been dequeued.
Overcoming this limitation would
require a refinement to matching schemes which can track additional relations
among data values.

Finally, in some cases it is unclear how to express matching schemes
deterministically. For example, each \emph{wait} operation of \emph{semaphore}
ADT should naturally match the \emph{notify} operation which enabled it.
However, it is not possible to determine this based on operation labels alone,
and it is not clear whether all implementations effectively keep track of this
correspondence. Choosing matches arbitrarily would be unsound, and
result in inferring ADTs with artificial constraints, e.g.,~that the notify and
wait operations behave according to a FIFO discipline, each wait matching the
oldest unmatched notify. Overcoming this limitation may require introducing
nondeterminism into matching schemes.

### 2015-11-02 histories.tex
Paragraph formatting.

\end{itemize}
The relation $<$ is an interval order because executions’ call and return
actions are totally ordered~\cite{conf/popl/BouajjaniEEH15}. Non-completed
operations are \emph{pending}, and are maximal in the happens-before order.

@@

The \emph{width} of a history is the maximum number of its operations which are
mutually unordered, and the \emph{width} of a history set is the maximum width
of its elements. Width-$1$ histories are \emph{sequential}. Since the operation
identifiers of a history have no intrinsic meaning, we consider equality between
histories up to renaming of operation identifiers.

>>>>

\end{itemize}
The relation $<$ is an interval order since call and return
actions are totally ordered~\cite{conf/popl/BouajjaniEEH15}. Non-completed
operations are \emph{pending}, and are maximal in the happens-before order.

@@

The \emph{width} of a history is its maximum number of mutually-unordered
operations; the \emph{width} of a history set is the maximum width
of its elements. Width-$1$ histories are \emph{sequential}. Since operation
identifiers have no intrinsic meaning, we consider equality between
histories up to their renaming.

### 2015-11-02 intro.tex
Quick typo fix.
Removed kernel talk.
Paragraph formatting.

  \item Does every implementation execution have a valid linearization?

  \item Does every valid linearizations preserve client invariants?

@@

The fundamental challenge is in identifying the algebraic properties of ADTs
which allow us to characterize an infinite set of violating executions with a
finite set of patterns. This characterization is non-trivial since an execution
with more operations than a given violating execution is not necessarily a
violation itself. For instance, an execution which contains only a single
pop operation returning the value $1$ is a violation to the atomic stack
ADT, whereas an execution containing an additional push$(1)$ operation,
overlapping in time with the pop, is not. Further complication arises
from the infinite set of possible data values, i.e.,~method argument and return
values. Our patterns must be sensitive to the \emph{relation} among data values
without being sensitive to the data values themselves. For instance, a
sequential execution in which $1$ and $2$ are pushed and subsequently
popped in the same order violates the atomic stack ADT. Yet, while replacing
both values $1$ and $2$ with the value $1$ results in a
\emph{valid} stack execution, replacing them with $3$ and $4$ results
in a violation.

@@

\end{itemize}
We conclude with a discussion about the limitations of our approach
(§\ref{sec:discussion}) and related work (§\ref{sec:related}).

>>>>

  \item Is there a valid linearization for each implementation execution?

  \item Does every valid linearization preserve client invariants?

@@

The fundamental challenge is in identifying the algebraic properties of ADTs
which characterize an infinite set of violating executions with a
finite set of patterns. This characterization is non-trivial since an execution
with more operations than a given violating execution is not necessarily a
violation itself. For instance, an execution which contains only a single
pop operation returning the value $1$ is a violation to the atomic stack
ADT, whereas an execution containing an additional push$(1)$ operation,
overlapping in time with the pop, is not. Further complication arises
from the infinite set of possible data values, i.e.,~method argument and return
values. Our patterns must be sensitive to the \emph{relation} among data values
without being sensitive to the data values themselves. For instance, a
sequential execution in which $1$ and $2$ are pushed and subsequently
popped in the same order violates the atomic stack ADT. Yet, while replacing
both values $1$ and $2$ with the value $1$ results in a
\emph{valid} stack execution, replacing them with $3$ and $4$ results
in a violation.

@@

\end{itemize}
We conclude by discussing limitations
(§\ref{sec:discussion}) and related work (§\ref{sec:related}).

### 2015-11-02 nature.tex
Margins.
Paragraph formatting.

%!TEX root = draft.tex
\section{Naturally-Occurring ADTs}
\label{sec:nature}

In this section we demonstrate that the premises used in the development of our
symbolic ADT inference algorithm — i.e.,~uniqueness and bounded-width,
faithful and normalizing matching schemes, and predictability — hold for the
ADTs which are typically provided by concurrent object libraries. Furthermore,
we demonstrate that the algorithm developed in this work computes precise
symbolic representations for these ADTs which can be used in symbolic checkers
for observational refinement~\cite{conf/pldi/EmmiEH15}. Our publicly available
implementation\footnote{The URL has been suppressed for anonymity purposes.}
enumerates the sequential histories\footnote{Our current implementation is
limited to atomic ADTs. For non-atomic ADTs, the enumeration must cover all
$k$-width histories, for some $k \in \mathbb{N}$.} not admitted by reference
implementations of the undermentioned ADTs, keeping only the histories which
are not generated by any other. In each case, our algorithm terminates in a
matter of seconds.

>>>>

\section{Naturally-Occurring ADTs}
\label{sec:nature}

In this section we demonstrate that the premises used in the development of our
symbolic ADT inference algorithm — i.e.,~uniqueness and bounded-width,
faithful and normalizing matching schemes, and predictability — hold for the
ADTs which are typically provided by concurrent object libraries. Furthermore,
we demonstrate that the algorithm developed in this work computes precise
symbolic representations for these ADTs which can be used in symbolic checkers
for observational refinement~\cite{conf/pldi/EmmiEH15}. Our publicly available
implementation\footnote{\url{https://github.com/imdea-software/violin}}
enumerates the sequential histories\footnote{Our current implementation is
limited to atomic ADTs. For non-atomic ADTs, the enumeration must cover all
$k$-width histories, for some $k \in \mathbb{N}$.} not admitted by reference
implementations of the undermentioned ADTs, keeping only the histories which
are not generated by any other. In each case, our algorithm terminates in a
matter of seconds with a list of human-readable ADT violation patterns.

### 2015-11-02 overview.tex
Paragraph formatting.

\begin{figure}[t]
  \begin{minipage}[b]{0.49\linewidth}
    \begin{verbatim}
[1:X] remove => 1  #
---
[1:1] add(1)           #
[2:2] remove => empty    #
---
[1:2] remove => 1  #
[2:2] add(1)         #
---
[1:1] add(1)       #
[2:2] add(2)         #
[3:2] remove => 2      #
---
[1:1] add(1)       #
[2:1] remove => 1    #
[3:1] remove => 1      #
---
[1:1] add(1)           #
[2:2] remove => empty    #
[3:1] remove => 1          #
---
[1:1] add(1)       #
[2:2] add(2)         #
[3:2] remove => 2      #
[4:1] remove => 1        #
    \end{verbatim}
  \end{minipage}
  \hfill
  \begin{minipage}[b]{0.49\linewidth}
    \begin{verbatim}
[1:1] add(1)       #
[2:X] remove => 2    #
---
[1:X] remove => 1  #
[2:X] remove => 1    #
---
[1:X] remove => 1      #
[2:2] remove => empty    #
---
[1:1] remove => empty  #
[2:X] remove => 1        #
---
[1:1] add(1)       #
[2:2] add(2)         #
[3:X] remove => 3      #
---
[1:1] add(1)           #
[2:2] add(2)             #
[3:3] remove => empty      #
---
[1:1] add(1)       #
[2:1] remove => 1    #
[3:X] remove => 2      #
    \end{verbatim}
  \end{minipage}
  \caption{Invalid sequences according to the atomic queue ADT. The histories
  on the right are each redundant with some history on the left, and those on
  the left constitute a complete set. The “\#” symbols depict the time intervals
  spanned by operations horizontally, and the “{\tt [$i$:$j$]}” notation refers
  to the current line’s operation identifier $i$, and the identifier $j$ of
  its matching operation. Adds and empty removes match themselves, and the
  “{\tt X}” symbol denotes an absent match.}
  \label{fig:patterns}
\end{figure}

The key technical obstacle in realizing the abstract algorithm of
Algorithm~\ref{alg:abstract}, overcome in Section~\ref{sec:patterns}, is the
classification of violations into a finite set of patterns, rendering the
remaining invalid sequences redundant. The symbolic representation which
excludes a given set of patterns is relatively straightforward to construct, and
is given in Section~\ref{sec:formula}. Note however that the termination of this
abstract algorithm is nondeterministic, and thus the inferred ADT specification
is generally not the strongest possible in the sense that it may not exclude
certain invalid linearizations. Nevertheless, specifications inferred by this
algorithm are sound, in the sense that when modular program reasoning with
inferred specifications succeeds, correctness follows. In any case, in
Section~\ref{sec:algorithm} we give conditions under which our refinement
of this abstract algorithm is both sound and complete, and show that these
conditions hold for naturally-occurring ADTs, thus resulting in sound and
complete symbolic ADT specifications. The remainder of this article develops
the technical machinery required to realize this abstract inference algorithm.

>>>>

\begin{figure}[t]
  \begin{minipage}[b]{0.49\linewidth}
    \begin{verbatim}
[1:X] remove => 1  #
---
[1:1] add(1)           #
[2:2] remove => empty    #
---
[1:2] remove => 1  #
[2:2] add(1)         #
---
[1:1] add(1)       #
[2:2] add(2)         #
[3:2] remove => 2      #
---
[1:1] add(1)       #
[2:1] remove => 1    #
[3:1] remove => 1      #
---
[1:1] add(1)           #
[2:2] remove => empty    #
[3:1] remove => 1          #
---
[1:1] add(1)       #
[2:2] add(2)         #
[3:2] remove => 2      #
[4:1] remove => 1        #
    \end{verbatim}
  \end{minipage}
  \hfill
  \begin{minipage}[b]{0.49\linewidth}
    \begin{verbatim}
[1:1] add(1)       #
[2:X] remove => 2    #
---
[1:X] remove => 1  #
[2:X] remove => 1    #
---
[1:X] remove => 1      #
[2:2] remove => empty    #
---
[1:1] remove => empty  #
[2:X] remove => 1        #
---
[1:1] add(1)       #
[2:2] add(2)         #
[3:X] remove => 3      #
---
[1:1] add(1)           #
[2:2] add(2)             #
[3:3] remove => empty      #
---
[1:1] add(1)       #
[2:1] remove => 1    #
[3:X] remove => 2      #
    \end{verbatim}
  \end{minipage}
  \caption{Invalid sequences according to the atomic queue ADT. Histories
  on the right are each redundant with some on the left; those on
  the left constitute a complete set. The “\#” symbols depict the time intervals
  spanned by operations horizontally, and the “{\tt [$i$:$j$]}” notation refers
  to the current line’s operation identifier $i$, and the identifier $j$ of
  its matching operation. Adds and empty removes match themselves, and the
  “{\tt X}” symbol denotes an absent match.}
  \label{fig:patterns}
\end{figure}

The key technical obstacle in realizing Algorithm~\ref{alg:abstract},
which we overcome in Section~\ref{sec:patterns}, is
classifying violations into a finite set of patterns, rendering the
remaining invalid sequences redundant. The symbolic representation which
excludes a given set of patterns is relatively straightforward to construct, and
is given in Section~\ref{sec:formula}. Note however that the termination of this
abstract algorithm is nondeterministic, and thus the inferred ADT specification
is generally not the strongest possible in the sense that it may not exclude
certain invalid linearizations. Nevertheless, specifications inferred by this
algorithm are sound, in the sense that when modular program reasoning with
inferred specifications succeeds, correctness follows. In any case, in
Section~\ref{sec:algorithm} we give conditions under which our refinement
of this abstract algorithm is both sound and complete, and show that these
conditions hold for naturally-occurring ADTs, thus resulting in sound and
complete symbolic ADT specifications. The remainder of this article develops
the technical machinery required to realize this abstract inference algorithm.
